Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-08-03 writing/DraftExposurePaper/DraftExposurePaper.Rmd
resolve merge conflict

Ike wind data:

```{r}
#Ike-2008 wind
load("MWR_gauge_wind_data/ike_wind.RData")
ike_wind <- ike_wind %>%
        filter(!is.na(max_sust_kt)) %>%
        mutate(max_sust_mps = convert_wind_speed(max_sust_kt, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

ike_wind_model <- county_wind(counties = unique(ike_wind$fips),
                              start_year = 2008, end_year = 2008,
                              wind_limit = 0) %>%
        filter(storm_id == "Ike-2008")

compare_wind <- right_join(ike_wind_model, ike_wind, by = "fips")
a <- ggplot(compare_wind, aes(x = max_sust, y = max_sust_mps,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Ike, 2008")
a

county_range <- compare_wind %>% 
        group_by(fips) %>%
        summarize(lowest_wind = min(max_sust_mps),
                  highest_wind = max(max_sust_mps),
                  model_wind = mean(max_sust))
b <- ggplot(compare_wind, aes(x = max_sust, y = max_sust_mps)) + 
        geom_segment(data = county_range, 
                     aes(x = model_wind, xend = model_wind,
                         y = lowest_wind, yend = highest_wind,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_wind$max_sust, compare_wind$max_sust_mps)) +
        ylim(range(compare_wind$max_sust, compare_wind$max_sust_mps)) + 
        ggtitle("Ike, 2008") + 
        xlab("Modeled wind speed (m / s)") + 
        ylab("MWR wind reports (m / s)")
b
```

Frances wind data:

```{r}
load("MWR_gauge_wind_data/frances_wind.RData")
frances_wind <- frances_wind %>%
        filter(!is.na(wind_sust)) %>%
        mutate(max_sust_mps = convert_wind_speed(wind_sust, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

frances_wind_model <- county_wind(counties = unique(frances_wind$fips),
                              start_year = 2004, end_year = 2004,
                              wind_limit = 0) %>%
        filter(storm_id == "Frances-2004")

compare_wind <- right_join(frances_wind_model, frances_wind, by = "fips")
a <- ggplot(compare_wind, aes(x = max_sust, y = max_sust_mps,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Frances, 2004")
a

county_range <- compare_wind %>% 
        group_by(fips) %>%
        summarize(lowest_wind = min(max_sust_mps),
                  highest_wind = max(max_sust_mps),
                  model_wind = mean(max_sust))
b <- ggplot(compare_wind, aes(x = max_sust, y = max_sust_mps)) + 
        geom_segment(data = county_range, 
                     aes(x = model_wind, xend = model_wind,
                         y = lowest_wind, yend = highest_wind,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_wind$max_sust, compare_wind$max_sust_mps)) +
        ylim(range(compare_wind$max_sust, compare_wind$max_sust_mps)) + 
        ggtitle("Frances, 2004") + 
        xlab("Modeled wind speed (m / s)") + 
        ylab("MWR wind reports (m / s)")
b
```

@@

#### Are exposure patterns different when using distance-, rain-, and wind-based metrics? In particular, (1) do some exposure metrics result in more "exposed" classifications than others and (2) are geographic patterns of exposure different among the different metrics?

@@

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Ivan-2004"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Ivan, 2004")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Ivan, 2004")
grid.arrange(a, b, ncol = 1)
```

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Isidore-2002"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Isidore, 2002")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_event_exposure(storm = storm, 
                                           event_type = "flood")) + 
        ggtitle("Flood events during Isidore, 2002")
grid.arrange(a, b, ncol = 1)
```

@@

@@

```{r echo = FALSE, fig.width = 12, fig.height = 12}
a <- agreement_hist(all_exposed, "distance", "rain")
b <- agreement_hist(all_exposed, "distance", "wind")
c <- agreement_hist(all_exposed, "distance", "flood")
d <- agreement_hist(all_exposed, "distance", "tornado")
e <- agreement_hist(all_exposed, "distance", "wind_event")

>>>>

For four example storms, we also assessed how estimated wind speeds from our model compared to observed wind speeds reported in the *Monthy Weather Review* tropical cyclone reports (Figure \ref{fig:mwr_wind_comparison}).

```{r mwr_wind_comparison, fig.width = 10, fig.height = 10, echo = FALSE, fig.align = "center", fig.cap = "Comparison of estimated storm sustained winds from our method (using a model based on Willoughby's model) with ground-based observations reported in the Monthly Weather Review report. Observations within the same county are linked by a line segment. The diagonal gray line shows where points would fall if assessments were identical. Horizontal and vertical gray lines show the 15 m / s threshold we use for wind-related storm exposure classification."}
# Function to plot comparison between two datasets
plot_mwr_compare_wind <- function(data){
        storm_year <- as.numeric(gsub(".+-", "", data$storm_id[1]))
        data$fips <- sprintf("%05s", data$fips)
        storm_title <- paste(unlist(strsplit(as.character(data$storm_id[1]), 
                                       split = "-")), collapse = ", ")

        compare_wind <- county_wind(counties = unique(data$fips),
                             start_year = storm_year, 
                             end_year = storm_year,
                             wind_limit = 0,
                             wind_var = "max_sust") %>%
                dplyr::filter(storm_id == data$storm_id[1]) %>%
                dplyr::right_join(data, by = "fips")

        county_range <- compare_wind %>% 
                group_by(fips) %>%
                summarize(lowest_wind = min(wind_sust),
                  highest_wind = max(wind_sust),
                  model_wind = mean(max_sust))

        plot_mwr <- ggplot(compare_wind, 
                   aes(x = max_sust, y = wind_sust)) + 
        geom_hline(yintercept = 15, color = "lightgray") + 
        geom_vline(xintercept = 15, color = "lightgray") + 
                geom_abline(aes(intercept = 0, slope = 1), color = "gray") + 
        geom_segment(data = county_range, 
                     aes(x = model_wind, xend = model_wind,
                         y = lowest_wind, yend = highest_wind,
                         group = fips)) + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        xlim(range(compare_wind$max_sust, compare_wind$wind_sust)) +
        ylim(range(compare_wind$max_sust, compare_wind$wind_sust)) + 
        ggtitle(storm_title) + 
        xlab("Modeled sustained wind (m / s)") + 
        ylab("Monthly Weather Report\n sustained wind reports (m / s)") + 
        scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black"), 
                           guide = "none") 
        return(plot_mwr)
}

#allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv") 
#allison_mwr <- plot_mwr_compare(allison_gauge)

load("MWR_gauge_wind_data/frances_wind.RData")
frances_wind <- frances_wind %>%
        filter(!is.na(wind_sust)) %>%
        mutate(wind_sust = convert_wind_speed(wind_sust, "knots", "mps"))
frances_wind_mwr <- plot_mwr_compare_wind(frances_wind)
```

@@

```{r}
load("MWR_gauge_wind_data/ike_wind.RData")
ike_wind <- ike_wind %>%
        filter(!is.na(max_sust_kt)) %>%
        mutate(wind_sust = convert_wind_speed(max_sust_kt, "knots", "mps"))
ike_wind_mwr <- plot_mwr_compare_wind(ike_wind)

# load("MWR_gauge_rain_data/alex_gauge.Rdata")
# alex_mwr <- plot_mwr_compare(alex_gauge)

# grid.arrange(allison_mwr, frances_mwr, ike_mwr, alex_mwr, ncol = 2)
grid.arrange(frances_wind_mwr, ike_wind_mwr,
             frances_wind_mwr, ike_wind_mwr, ncol = 2)
```

#### Are exposure patterns different when using different exposure metrics? In particular, (1) do some exposure metrics result in more "exposed" classifications than others and (2) are geographic patterns of exposure different among the different metrics?

We determined the average number of hits per year for each eastern US county for the exposure criteria based on distance (Figure \ref{fig:distance_hits}), rain (Figure \ref{fig:rain_hits}), and wind (Figure \ref{fig:wind_hits}), as well as the total number of hits based on storm-associated events listed in the NOAA Storm Events database for flood events (Figure \ref{fig:flood_hits}), wind events (Figure \ref{fig:wind_event_hits}), and tornado events (Figure \ref{fig:tornado_hits}) (all of these maps based on NOAA Storm Events are restricted to storms between 1996 and 2011).

```{r echo = FALSE}
# Map number of hits per county based on exposure data
map_total_hits <- function(exposure_data, my_fips = my_fips, 
                           eastern_states = eastern_states,
                           title, start_year = 1988, end_year = 2011){
        missing_fips <- my_fips[!(my_fips %in% exposure_data$fips)]
        missing_counties <- data.frame(region = as.numeric(missing_fips),
                                       count = 0)

        exposure_count <- exposure_data %>%
                dplyr::group_by(fips) %>%
                dplyr::summarize(count = n() /(end_year - start_year + 1)) %>%
                dplyr::rename(region = fips) %>%
                dplyr::mutate(region = as.numeric(region),
                              count = as.numeric(count)) %>%
                rbind(missing_counties)

        exposure_count$value <- cut(exposure_count$count, c(0, 
                                                            0.11,
                                                            0.22,
                                                            0.33,
                                                            0.44,
                                                            0.56,
                                                            0.67,
                                                            0.78,
                                                            0.89,
                                                            10),
                           labels = c("0-.10", ".11-.21",
                                      ".22-.32", ".33-.43",
                                      ".44-.55", ".56-.66",
                                      ".67-.77", ".78-.88", ".89+"),
                           include.lowest = TRUE)
        exposure_palette <- c("#FFFFFF", brewer.pal(8, name = "YlGnBu"))

        map_dist_count <- CountyChoropleth$new(exposure_count)
        map_dist_count$set_zoom(eastern_states)
        map_dist_count$ggplot_scale <- scale_fill_manual(
                name = "Hits per\ncounty per\nyear",
                values = exposure_palette,
                na.value = "white")
        map_dist_count$title <- title
        a <- map_dist_count$render()

        storms_to_plot <- exposure_data %>%
                dplyr::group_by(storm_id) %>%
                dplyr::summarize(n = n()) %>%
                dplyr::arrange(desc(n)) %>%
                slice(1:5)
        # print(storms_to_plot)
        out <- map_tracks(storms_to_plot$storm_id, plot_object = a,
                          alpha = 0.75, plot_points = FALSE)
        return(out)
}
```

```{r distance_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on a criterion of the county being within 100 kilometers of the storm track, for storms from 1988 to 2011. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Beryl, 1994: 346 counties; Frances, 2004: 321 counties; Cindy, 2005: 317 counties; Jeanne, 2004: 296 counties; and Dennis, 2005: 290 counties)."}
distance_exposure <- county_distance(counties = my_fips, start_year = 1988,
                                     end_year = 2011, dist_limit = 100)
map_total_hits(distance_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on distance criterion")
```

```{r rain_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on a criteria of the county getting 75 millimeters or more of storm-associated rain in the window from two days before to one day after the storm's closest approach to the county and the county being within 500 kilometers of the storm track, for storms from 1988 to 2011. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criteria over the same period (Frances, 2004: 464 counties; Ivan, 2004, 411 counties; Opal, 1995: 407 counties; Ike, 2008: 338 counties, Lee, 2011: 355 counties)."}
rain_exposure <- county_rain(counties = my_fips,
                             start_year = 1988, end_year = 2011,
                             rain_limit = 75, dist_limit = 500)
map_total_hits(rain_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on rain criterion")
```

```{r wind_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on a criteria of the county experiencing storm-related sustained gusts of 15 m / s or more, for storms from 1988 to 2011. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Bertha, 1996: 199 counties; Irene, 2011: 139 counties; Floyd, 1999: 135 counties; Isabel, 2003: 128 counties; Bob, 1991: 118 counties)."}
wind_exposure <- county_wind(counties = my_fips, start_year = 1988,
                             end_year = 2011, wind_limit = 15)
map_total_hits(wind_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on wind criterion")
```

```{r flood_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on the county having a storm-associated flood event listed in the NOAA Storm Events database, for storms from 1996 to 2011. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Floyd, 1999: 218 counties; Allison, 2001: 187 counties; Irene, 2011: 187 counties; Ike, 2008: 167 counties; Ivan, 2004: 153 counties)."}
flood_exposure <- county_events(counties = my_fips,
                              start_year = 1996, end_year = 2011,
                              event_type = "flood")
map_total_hits(flood_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on flood events",
               start_year = 1996, end_year = 2011)
```

```{r wind_event_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on the county having a storm-associated wind event listed in the NOAA Storm Events database, for storms from 1996 to 2011. Light red lines show the storm tracks over the same period. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Allison, 2001: 435 counties; Arlene, 2005: 360 counties; Danny, 1997: 310 counties; Cindy, 2005: 189 counties; Cristobal, 2008: 183 counties)."}
wind_event_exposure <- county_events(counties = my_fips,
                              start_year = 1996, end_year = 2011,
                              event_type = "wind")
map_total_hits(wind_event_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on wind events")
```

```{r tornado_hits, echo = FALSE, fig.width = 8, fig.height = 6, fig.align = "center", fig.cap = "Average number of hits per year for eastern US counties based on the county having a storm-associated tornado event listed in the NOAA Storm Events database, for storms from 1996 to 2011. Light red lines show the storm tracks over the same period. Red lines show the storm tracks for the top 5 storms in terms of number of counties classified as 'exposed' to the storm based on this criterion over the same period (Ivan, 2004: 95 counties; Frances, 2004: 72 counties; Rita, 2005: 52 counties; Katrina, 2005: 45 counties; Fay, 2008: 43 counties)."}
tornado_exposure <- county_events(counties = my_fips,
                              start_year = 1996, end_year = 2011,
                              event_type = "tornado")
map_total_hits(tornado_exposure, my_fips = my_fips, 
               eastern_states = eastern_states,
               title = "Hits per county based on tornado events",
               start_year = 1996, end_year = 2011)
```

@@

@@

```{r echo = FALSE, eval = FALSE}
all_storm_counties <- closest_dist %>%
        dplyr::select(storm_id, fips) %>%
        dplyr::mutate(storm_county = paste(storm_id, fips, sep = "-")) %>%
        dplyr::filter(as.numeric(gsub(".*-", "", storm_id)) >= 1996 & 
                              as.numeric(gsub(".*-", "", storm_id)) <= 2011)

@@

```{r echo = FALSE, fig.width = 12, fig.height = 12, eval = FALSE}
a <- agreement_hist(all_exposed, "distance", "rain")
b <- agreement_hist(all_exposed, "distance", "wind")
c <- agreement_hist(all_exposed, "distance", "flood")
d <- agreement_hist(all_exposed, "distance", "tornado")
e <- agreement_hist(all_exposed, "distance", "wind_event")
